---
title: "Weekly 2: Coffee Ratings"
author: "[Name]"
format:
  html:
    embed-resources: true
editor: visual
---

## Overview

You want to know which producers grow the best-rated coffee.

That sounds like a `group_by()` and a `summarise()`. The problem is that `producer` was typed by different people in different ways, so the same producer can appear under several labels, and some labels are not producers at all.

Your job is to work out how much of that you can fix, how much you shouldn't, and what that does to the answer.

A few things:

1.  Add `#| echo: false` to your code blocks.
2.  Use a new block when you move to a new task.
3.  Code goes in the blocks. **Answers go outside them**, as prose. A comment inside a code chunk is not an answer.
4.  Keep the original `producer` column. Every cleaned version goes in a **new** column.

Due Monday at 8pm.

## Your data

Import `coffee_ratings.csv` and inspect it.

```{r}
#| warning: false
library(tidyverse)
coffee_ratings <- read_csv("data/raw/coffee_ratings.csv")

glimpse(coffee_ratings)
```

## Before you start

> What does one row represent?
>
> One row represents a sample that had been graded and information about its origin.

> Is there a column that uniquely identifies a row? How would you check?
>
> No. Compare n_row and n_distinct

> Is `producer` a key? What does it mean that the same producer can appear on many rows?
>
> No. It means that one produver can have multiple samples in the dataset.

## How many producers?

Count the distinct values of `producer`.

```{r}
#| warning: false

coffee_ratings |> 
  count(producer, sort = TRUE)

n_distinct(coffee_ratings$producer)
sum(is.na(coffee_ratings$producer))
n_distinct(coffee_ratings$producer, na.rm = FALSE)
```

> How many distinct values are there?
>
> 692

> How many rows have no producer recorded? Does your distinct count include "missing" as one of its values? How would you check?
>
> 232. No. Compare the 3 lines I had at the bottom.
>
> Why shouldn't you read that number as the number of distinct producers?
>
> The same producer can be written several different ways, we can not expect every person to insert data the exact same way.

## Normalize the representation

Create a new column, `producer_clean`, that makes the producer labels consistent in case and whitespace. Keep `producer`.

```{r}
#| warning: false
coffee_ratings_clean <- coffee_ratings |> 
  mutate(
    producer_clean = producer |> 
      str_to_lower() |> 
      str_squish()
  )

n_distinct(coffee_ratings_clean$producer_clean)
```

> How many distinct values does `producer_clean` have?
>
> 666

> Is that the number of producers now? What does the reduction establish, and what does it *not* establish?
>
> No. It shows that 26 case and space differences were removed, it does not show us that there are exactly 666 producers.

## Which labels actually merged?

A single number doesn't tell you what happened. Build a table showing every `producer_clean` value that more than one raw `producer` value was combined into, along with the raw values that went into it.

```{r}
#| warning: false
coffee_ratings_clean |> 
  filter(!is.na(producer_clean)) |> 
  group_by(producer_clean) |> 
  summarise(
    n_raw_labels = n_distinct(producer),
    raw_values = paste(sort(unique(producer)), collapse = " | ")
  ) |> 
  filter(n_raw_labels > 1) 
```

> How many cleaned labels absorbed more than one raw label?
>
> 22

> Look through them. Is every merge one you would stand behind?
>
> No, I saw varios and various as seperate ones.

> Is this table more useful than knowing the number of labels went down? Why?
>
> Yes. It shows what change.

## Does it change anything?

Compare the La Plata coffees before and after normalizing. For each version, report the number of coffees and the mean and median `total_cup_points`.

```{r}
#| warning: false
la_plata <- coffee_ratings |> 
  filter(producer == "La Plata") |> 
  summarise(
    n_coffees = n(),
    mean_score = mean(total_cup_points, na.rm = TRUE),
    median_score = median(total_cup_points, na.rm = TRUE)
  )

la_plata_clean <- coffee_ratings_clean |> 
  filter(producer_clean == "la plata") |> 
  summarise(
    n_coffees = n(),
    mean_score = mean(total_cup_points, na.rm = TRUE),
    median_score = median(total_cup_points, na.rm = TRUE)
  )

bind_rows(la_plata, la_plata_clean)
```

> What changed more: the number of coffees or the average score?
>
> Number of coffees changed from 30 to 39 and the median median scores also changed.

> Suppose you had reported how many coffees La Plata produced using the raw column. How far off would you have been?
>
> 9

## What case and whitespace didn't catch

Some labels still differ only in punctuation, i.e., a period, a comma, a slash.

Sort the distinct values of `producer_clean` and read through them. Near-duplicates end up next to each other.

```{r}
#| warning: false
coffee_ratings_clean |> 
  select(producer_clean) |> 
  distinct() |> 
  filter(!is.na(producer_clean)) |> 
  arrange(producer_clean) 
```

> Find at least **three** pairs of labels that differ only in punctuation.
>
> kerchanshe trading p.l.c and kerchanshe trading plc
>
> yeh, li chuan and yeah, li chuan.
>
> conquista / morito and conquista/ morito

Now use exact replacement to make each pair consistent. Change only the values you identified and nothing else.

```{r}
#| warning: false
coffee_ratings_clean <- coffee_ratings_clean |> 
  mutate(
    producer_clean_punctuation = case_when(
      producer_clean == "kerchanshe trading p.l.c" ~ "kerchanshe trading plc",
      producer_clean == "yeh, li chuan." ~ "yeh, li chuan",
      producer_clean == "conquista / morito" ~ "conquista/ morito"
    )
  )

coffee_ratings_clean |> 
  filter(producer_clean != producer_clean_punctuation) |> 
  count(producer_clean, producer_clean_punctuation)
```

> For each change, say why you're confident it's the same producer.
>
> For conquista it is a single space difference, a lot of people do not use a space before using a / while some do.
>
> For kerchanshe plc seems to be an acronym so it makes sense that some people will punctuate while others wouldnt.
>
> For yeh it makes sense that someone would put a period at the end out of habit.

> How long did it take to find those by reading? How many labels did you have to look at?
>
> It took SEVERAL minutes I had to look through almost all of the dataset.

## The top producers

Using `producer_clean`, find the ten producers with the most coffees in the data.

```{r}
#| warning: false
top_ten_producers <- coffee_ratings_clean |> 
  filter(!is.na(producer_clean)) |> 
  count(producer_clean, sort = TRUE) |> 
  head(n = 10)

top_ten_producers
```

Look at that list carefully.

> Is every entry in it a producer? If not, which ones aren't, and what do they mean?
>
> No, varios means various.

> Is any single producer split across more than one row of the list?
>
> Yes. ipanema agricola sa, ipanema agricola, ipanema agricola s.a

Now look at these four raw labels:

- Ipanema Agricola
- Ipanema Agricola S.A
- Ipanema Agrícola SA
- Ipanema Coffees

> Should these be combined into one producer? **Don't recode them.** Explain whether combining them would be normalization or classification, and what evidence you'd need before doing it.
>
> The middle two should be combined, however we do not have enoug information to determine the others. These could all share a parent company but be completely different entities or it could be a region that all of the companies named themselves after.

> If you did combine them, what would happen to your top-ten list?
>
> It would combine 4 into one so that one would become a lot larger and 3 more companies would get moved up onto the list.

## What this means for your question

> A classmate says: "I reduced the producer column from 692 labels to 666, so the cleaned column is better." Is that justified? Explain.
>
> Yes but that does not mean it is perfect. They likely combined some companies that does not mean it got repeat in the set.

> You set out to find which producers grow the best-rated coffee. Given everything above, what would you need to tell someone before they trusted your answer?
>
> How I cleaned the labels, which possible matches I did not combine or did combine, and how I handled lables such as varios.

## Disclosure appendix

> What did you try first? Include the code you wrote and the actual error message, or a clear description of the problem if there was no error.

> If you used an AI system: the exact prompt, exactly what it returned, what it got wrong or left unclear, and what you changed.

> If you didn't use one, say so.

Original:

coffee_ratings_clean \|\> filter(!is.na(producer_clean)) \|\> group_by(producer_clean) \|\> summarise( n_raw_labels = n_distinct(producer) \|\> raw_values = paste(sort(unique(producer)), collapse = " \| ") ) \|\> filter(n_raw_labels \> 1)

Note: I had to use google to find out to add paste(sort(unique(producer)), collapse = " \| ") )\
Error: unexpected '=' in: " n_raw_labels = n_distinct(producer) \|\> raw_values ="

Asked chatgpt:

Why isn't the following code working coffee_ratings_clean \|\> filter(!is.na(producer_clean)) \|\> group_by(producer_clean) \|\> summarise( n_raw_labels = n_distinct(producer) \|\> raw_values = paste(sort(unique(producer)), collapse = " \| ") ) \|\> filter(n_raw_labels \> 1)

Chat's response:

The problem is a **syntax error inside `summarise()`**. You used the pipe `|>` between two separate summary variables.

The pipe should not be there. Each summary variable needs to be separated by a comma.

Use:

```         
coffee_ratings_clean |>   filter(!is.na(producer_clean)) |>   group_by(producer_clean) |>   summarise(     n_raw_labels = n_distinct(producer),     raw_values = paste(sort(unique(producer)), collapse = " | ")   ) |>   filter(n_raw_labels > 1)
```
